---
title: "01_HDP_kinfitr_preprocessing"
output: html_document
date: "2025-12-09"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Setup

## Packages

```{r}
# remotes::install_github("mathesong/bloodstream")
# remotes::install_github("mathesong/kinfitr")

rm(list = ls())  # Clear all objects from the workspace

library(tidyverse)
library(bloodstream)
library(kinfitr)
library(job)
library(dplyr)
library(tidyr)
library(readr)
library(ggpubr)
library(jsonlite)

theme_set(theme_light())

```

# Loading Data

## Loading IDIF and participants.tsv files

```{r}
suppressWarnings({

find_repo_root <- function(start = getwd()) {
  current <- normalizePath(start, winslash = "/", mustWork = TRUE)
  repeat {
    if (file.exists(file.path(current, "config", "paths.json"))) {
      return(current)
    }
    parent <- dirname(current)
    if (identical(parent, current)) {
      stop("Could not locate repo root containing config/paths.json")
    }
    current <- parent
  }
}
repo_root <- find_repo_root(getwd())
source(file.path(repo_root, "code", "utils", "project_paths.R"))
paths <- load_project_paths(repo_root)
bids_dir <- paths$bids_root
kinfitr_output_dir <- file.path(paths$kinfitr_output_root)

# Define the paths relative to the base path
idif_path <- file.path(bids_dir, "derivatives/idif_aorta")
participants_file <- file.path(bids_dir, "participants.tsv")

# Read the participants.tsv file
participants <- read.delim(participants_file)

# Initialize an empty dataframe to store results
idif_data <- tibble(
  ses = character(),
  sub = character(),
  condition = character(), # Add the condition column
  task = character(),
  trc = character(),
  acq = character(),
  run = character(),
  rec = character(),
  input_idif = list(),
  input = list(), # this will be formatted as needed by kinfitr
  input_feng = list() # add the input function with Feng model
)

glc_data  <- tibble(
  ses = character(),
  sub = character(),
  condition = character(), # Add the condition column
  task = character(),
  trc = character(),
  acq = character(),
  run = character(),
  rec = character(),
  glc_conc = numeric()
)

# Traverse the derivatives folder to find the relevant files
files <- list.files(path = idif_path, pattern = "inputfunction\\.tsv", full.names = TRUE, recursive = TRUE)

# Process each file
for (file in files) {
  # Extract sub and ses from the filename
  filename <- basename(file)
  sub <- gsub("sub-(HDP[0-9]+)_.*", "\\1", filename)
  ses <- gsub(".*_ses-([0-9]+)_.*", "\\1", filename)
  
  print(paste("Processing participant", sub, "session", ses)) 
  
  # Read the original input function file
  input_data <- read.delim(file)
  
  # Add timepoint 0 to the dataframe
  zero_row <- input_data[1,] %>% mutate(across(everything(), ~0))
  input_data <- bind_rows(zero_row, input_data)
  
  # Extract original timepoints and frame info
  original_times <- input_data$time
  frame_info <- input_data %>% select(Frame, FrameTimesStart, FrameDuration)
  
  # Determine the drug condition based on the session
  condition <- ifelse(ses == "01", 
                      participants$drug_ses.01[participants$participant_id == paste0("sub-", sub)], 
                      participants$drug_ses.02[participants$participant_id == paste0("sub-", sub)])
  
  ### Calculate blood plasma concentration from population red blood cell to plasma ratio (Phelps et al. 1979)
  t_min <- input_data$time / 60   # Time in minutes
  rbc_plasma <- 0.8 + 0.0012 * t_min      # [RBC]/[plasma] ratio r(t) from Phelps
  H <- 0.42 # Hematocrit (otherwise assume 0.42)
  denom <- H * rbc_plasma + (1 - H) # Denominator H * r(t) + (1 - H)
  plasma_idif <- input_data$IDIF / denom   # Plasma TAC from whole blood

  
  # Use blood_interp() to create the input list based on input_idif data
  input <- blood_interp(
    t_blood      = t_min,
    blood        = input_data$IDIF,   # whole blood
    t_plasma     = t_min,
    plasma       = plasma_idif,       # corrected plasma
    t_parentfrac = t_min,
    parentfrac   = 1 # Assuming constant for FDG
  )

  # Fit the Feng model on whole-blood IDIF
  fengfit <- blmod_feng(input_data$time, input_data$IDIF)

  # Generate smooth predictions (whole-blood curve)
  fengpred <- tibble(time = seq(0, max(input_data$time), length.out = 1000))
  fengpred$IDIF <- predict(fengfit, newdata = fengpred)

  # Calculate plasma TAC from Feng whole-blood IDIF using the same RBC/plasma model
  t_min_feng <- fengpred$time / 60
  rbc_plasma_feng <- 0.8 + 0.0012 * t_min_feng
  denom_feng <- H * rbc_plasma_feng + (1 - H)
  plasma_idif_feng <- fengpred$IDIF / denom_feng

  # Convert to the required format for kinfitr
  input_feng <- blood_interp(
    t_blood      = t_min_feng,
    blood        = fengpred$IDIF,      # whole blood
    t_plasma     = t_min_feng,
    plasma       = plasma_idif_feng,   # corrected plasma
    t_parentfrac = t_min_feng,
    parentfrac   = 1
  )

  # Create a dataframe for Feng-modeled storage (Time remains in seconds here)
  feng_df <- tibble(
    Time           = fengpred$time,
    Blood          = fengpred$IDIF,        # whole blood
    Plasma         = plasma_idif_feng,     # corrected plasma
    ParentFraction = 1,
    AIF            = plasma_idif_feng      # plasma-based AIF
  )

  # Define the output directory
  output_dir <- dirname(file)
  feng_tsv_path <- file.path(output_dir, paste0("sub-", sub, "_ses-", ses, "_inputfunction_feng_modeled.tsv"))
  feng_json_path <- file.path(output_dir, paste0("sub-", sub, "_ses-", ses, "_inputfunction_feng_modeled.json"))

  # Save the Feng modeled data as TSV
  write.table(feng_df, file = feng_tsv_path, sep = "\t", row.names = FALSE, quote = FALSE)

  # Create BIDS-compliant JSON metadata
  feng_json <- list(
    Description = "Arterial Input function derived from original IDIF inputfunction.tsv created with blmod_feng function from kinfitr package.",
    Time = list(
      Description = "Reference time for the frame in relation to time zero defined in _pet.json",
      Units = "s"
    ),
    Blood = list(
      Description = "Radioactivity concentration in whole blood",
      Units = "Bq/mL"
    ),
    Plasma = list(
      Description = "Radioactivity concentration in plasma",
      Units = "Bq/mL"
    ),
    ParentFraction = list(
      Description = "Fraction of parent compound in plasma",
      Units = "unitless"
    ),
    AIF = list(
      Description = "Arterial input function derived from modeled blood data",
      Units = "Bq/mL"
    )
  )

  # Save JSON metadata
  write_json(feng_json, path = feng_json_path, pretty = TRUE)
  
  print(paste("Saved:", feng_tsv_path))

  ## ---- Store Back-Translated Feng Data (Original Format) ---- ##
  
  # Interpolate AIF values to match the original timepoints
  interpolated_aif <- approx(x = feng_df$Time, y = feng_df$AIF, xout = original_times, method = "linear")$y

  # Create a dataframe with original format
  feng_original_format <- tibble(
    time = original_times,
    IDIF = interpolated_aif,
    Frame = frame_info$Frame,
    FrameTimesStart = frame_info$FrameTimesStart,
    FrameDuration = frame_info$FrameDuration
  )
  
  # Remove the row where Frame == 0
  feng_original_format <- feng_original_format %>% filter(Frame != 0)

  
  # Define the output paths
  feng_original_tsv_path <- file.path(output_dir, paste0("sub-", sub, "_ses-", ses, "_inputfunction_feng_original_format.tsv"))
  feng_original_json_path <- file.path(output_dir, paste0("sub-", sub, "_ses-", ses, "_inputfunction_feng_original_format.json"))

  # Save the new TSV file
  write.table(feng_original_format, file = feng_original_tsv_path, sep = "\t", row.names = FALSE, quote = FALSE)

  # Create JSON metadata
  feng_original_json <- list(
      Description = "Arterial Input function derived from original IDIF inputfunction.tsv created with blmod_feng function from kinfitr package and reverted back to the original format found in inputfunction.tsv files.",
    time = list(
      Description = "Reference time for the frame in relation to time zero defined in _pet.json",
      Units = "s"
    ),
    IDIF = list(
      Description = "Inputfunction derived from IDIF and modeled with Feng model provided by kinfitr funtion 'blmod_feng'",
      Units = "Bq/mL"
    ),
    Frame = list(
      Description = "Frame index",
      Units = "unitless"
    ),
    FrameTimesStart = list(
      Description = "Time of frame start in relation to time zero defined in _pet.json",
      Units = "s"
    ),
    FrameDuration = list(
      Description = "Time of frame start in relation to time zero defined in _pet.json",
      Units = "s"
    )
  )

  # Save JSON metadata
  write_json(feng_original_json, path = feng_original_json_path, pretty = TRUE)

  print(paste("Saved:", feng_original_tsv_path))

  ## ---- Add to DataFrames for Further Processing ---- ##
  
  # Add a new row to the dataframe
  idif_data <- idif_data %>%
    add_row(
      ses = ses,
      sub = sub,
      condition = condition,
      task = "rest",
      trc = "trc",
      acq = "acq",
      run = "01",
      rec = "rec",
      input_idif = list(input_data),
      input = list(input),
      input_feng = list(input_feng)
    )

  # Extract the glucose concentration
  glc_conc <- ifelse(ses == "01", 
                     participants$blood_glucose_ses.01[participants$participant_id == paste0("sub-", sub)], 
                     participants$blood_glucose_ses.02[participants$participant_id == paste0("sub-", sub)])

  # Add glucose concentration to dataframe
  glc_data <- glc_data %>%
    add_row(
      ses = ses,
      sub = sub,
      condition = condition,
      task = "rest",
      trc = "trc",
      acq = "acq",
      run = "01",
      rec = "rec",
      glc_conc = glc_conc
    )
}

})
```

```{r}
head(idif_data, n = 6)

idif_data %>% 
  slice(1:6) %>% 
  pull(input_feng) %>% 
  map(plot)
```

## Loading petprep outputs (TACs and morphology)

```{r}
petprep_data <- kinfitr::bids_parse_files(file.path(bids_dir, "derivatives/petprep_extract_tacs_smoothed")) %>% 
  unnest(filedata) %>% 
  filter(str_detect(path_absolute, "gtmseg"))

tacs <- petprep_data %>% 
  filter(measurement=="tacs") %>% 
  filter(is.na(pvc)) %>% 
  mutate(tacdata = map(path_absolute, ~read_delim(.x, delim="\t", 
                                                  show_col_types = FALSE)))

morphdata <- petprep_data %>%
  filter(measurement=="morph") %>%
  mutate(morphdata = map(path_absolute, ~read_delim(.x, delim="\t",
                                                  show_col_types = FALSE)))
```


## Creating combined TACs

Now we want to combine regions by volume-weighted averages of the contitutent regions.  
So we combine the TACs and the region sizes

```{r}
selected_tacs <- select(tacs, c(ses:rec, tacdata)) %>% 
  inner_join(select(morphdata, c(ses:rec, morphdata))) %>% 
  group_by(sub, ses) %>% 
  mutate(tacdata = map(tacdata, ~pivot_longer(.x, 
                                              cols = `Left-Cerebral-White-Matter`:`ctx-rh-insula`, 
                                              names_to = "name", values_to = "TAC"))) %>% 
  mutate(tacdata = map2(tacdata, morphdata, ~inner_join(.x, .y, by="name")))
```

## Defining the regions of interest

```{r}

## Whole Cortex
regions_whole_ctx <-selected_tacs$morphdata[[1]] %>% 
  filter(str_detect(name, "ctx")) %>% 
  pull(name)

print("Cortex")
print(regions_whole_ctx)

## Left Cortex
regions_left_ctx <-selected_tacs$morphdata[[1]] %>% 
  filter(str_detect(name, "ctx-lh")) %>% 
  pull(name)

## Right Cortex
regions_right_ctx <-selected_tacs$morphdata[[1]] %>% 
  filter(str_detect(name, "ctx-rh")) %>% 
  pull(name)

##
exclude_areas <- c("CSF", "AirCavity", "Skull", "CSF-ExtraCerebral", "Head-ExtraCerebral", "Left-Cerebral-White-Matter", "Left-Cerebellum-White-Matter", "3rd-Ventricle", "4th-Ventricle", "Left-Lateral-Ventricle", "Left-Inf-Lat-Vent", "Right-Lateral-Ventricle", "Right-Inf-Lat-Vent", "Optic-Chiasm")
regions_whole_brain <- selected_tacs$morphdata[[1]] %>%
  filter(!str_detect(name, paste(exclude_areas, collapse = "|"))) %>%
  pull(name)
print("Whole Brain")
print(regions_whole_brain)

## Frontal Cortex
regions_frontal_ctx <- selected_tacs$morphdata[[1]] %>% 
  filter(str_detect(name, "frontal")) %>% 
  pull(name)

## Insula
regions_insula <- selected_tacs$morphdata[[1]] %>% 
  filter(str_detect(name, "insula")) %>% 
  pull(name)

## Striatum
regions_striatum <- selected_tacs$morphdata[[1]] %>% 
  filter(str_detect(name, "Putamen|Accumbens|Caudate")) %>% 
  pull(name)

## Hippocampus
regions_hippocampus <- selected_tacs$morphdata[[1]] %>% 
  filter(str_detect(name, "Hippocampus")) %>% 
  pull(name)

## Amygdala
regions_amyg <- selected_tacs$morphdata[[1]] %>% 
  filter(str_detect(name, "Amygdala")) %>% 
  pull(name)

## Thalamus
regions_thalamus <- selected_tacs$morphdata[[1]] %>% 
  filter(str_detect(name, "Thalamus")) %>% 
  pull(name)

## Caudate nucleus
regions_caudate_nucleus <- selected_tacs$morphdata[[1]] %>% 
  filter(str_detect(name, "Caudate")) %>% 
  pull(name)

## Putamen
regions_putamen <- selected_tacs$morphdata[[1]] %>% 
  filter(str_detect(name, "Putamen")) %>% 
  pull(name)

## Pallidum
regions_pallidum <- selected_tacs$morphdata[[1]] %>% 
  filter(str_detect(name, "Pallidum")) %>% 
  pull(name)

## Nucleus accumbens
regions_nucleus_accumbens <- selected_tacs$morphdata[[1]] %>% 
  filter(str_detect(name, "Accumbens")) %>% 
  pull(name)

## Choroid plexus
regions_choroid_plexus <- selected_tacs$morphdata[[1]] %>% 
  filter(str_detect(name, regex("choroid|plexus", ignore_case = TRUE))) %>% 
  pull(name)

## Ventral diencephalon
regions_ventral_diencephalon <- selected_tacs$morphdata[[1]] %>% 
  filter(str_detect(name, "VentralDC")) %>% 
  pull(name)

```

Then we perform volume-weighted averaging

```{r}
volume_weighted_average_tac <- function(tacdata, regions, regionname) {
  
  tacdata_combined <- tacdata %>% 
    filter(name %in% regions) %>% 
    group_by(frame_start, frame_end) %>% 
    mutate(volume_tot = sum(`volume-mm3`),
           volume_frac = `volume-mm3` / volume_tot,
           TAC_frac = TAC * volume_frac) %>% 
    summarise(!!regionname := sum(TAC_frac), 
              .groups = "keep") %>% 
    ungroup()
  
  return(tacdata_combined)
  
}

volume_weighted_average_tacs <- function(tacdata, regions_list) {
  regionnames <- names(regions_list)
  out <- map2(regions_list, regionnames, ~volume_weighted_average_tac(tacdata, .x, .y))
  out <- purrr::reduce(out, dplyr::inner_join, by = c("frame_start",
                                                      "frame_end"))
  return(out)
}

regions_list <- list(
  whole_cortex = regions_whole_ctx,
  left_cortex = regions_left_ctx,
  right_cortex = regions_right_ctx,
  whole_brain = regions_whole_brain,
  frontal_cortex = regions_frontal_ctx,
  insula = regions_insula,
  striatum = regions_striatum,
  hippocampus = regions_hippocampus,
  amygdala = regions_amyg,
  thalamus = regions_thalamus,
  caudate_nucleus = regions_caudate_nucleus,
  putamen = regions_putamen,
  pallidum = regions_pallidum,
  nucleus_accumbens = regions_nucleus_accumbens,
  ventral_diencephalon = regions_ventral_diencephalon
)

if (length(regions_choroid_plexus) > 0) {
  regions_list$choroid_plexus <- regions_choroid_plexus
}

regions_vector <- names(regions_list)

selected_tacs <- selected_tacs %>% 
  group_by(sub, ses) %>% 
  mutate(selected_tacdata = map(tacdata, ~volume_weighted_average_tacs(.x, 
                                                                       regions_list)))

## Combining the data
modeldata <- selected_tacs %>% 
  select(ses:rec, tacdata = selected_tacdata) %>% 
  inner_join(idif_data)

```


# Fitting TACs

## Preparation

```{r}

### Adding frame midtimes and durations
modeldata <- modeldata %>% 
  mutate(tacdata = map(tacdata, ~.x %>% 
                         mutate(frame_start = frame_start / 60,
                                frame_end = frame_end / 60) %>% 
                         mutate(frame_dur = frame_end - frame_start,
                                frame_mid = frame_start + 0.5*frame_dur)))


### Adding weights
modeldata <- modeldata %>% 
  mutate(tacdata = map(tacdata, ~.x %>% 
                         mutate(meanTAC = rowMeans( .x %>% select(whole_brain) )) %>% 
                         mutate(weights = weights_create(t_start = frame_start,
                                                         t_end = frame_end,
                                                         tac = meanTAC,
                                                         radioisotope = "F18"))))
```

### Plot the tacs for each region
```{r}
#---------------------------------------------------------------------
# 1) Per-scan / per-region TAC plots
#---------------------------------------------------------------------
for (i in seq_len(nrow(modeldata))) {
  sub  <- modeldata$sub[i]
  ses  <- modeldata$ses[i]
  cond <- modeldata$condition[i]

  # dynamic table for this scan (already has frame_mid)
  tac_df <- modeldata$tacdata[[i]]

  for (reg in regions_vector) {
    outdir <- file.path(kinfitr_output_dir, "02_Outcomes", "01_TAC_plots", reg)
    dir.create(outdir, recursive = TRUE, showWarnings = FALSE)

    fname   <- sprintf("sub-%s_ses-%s_%s_desc-%s_TAC.png",
                       sub, ses, cond, reg)
    outpath <- file.path(outdir, fname)

    png(outpath, width = 1000, height = 800)
    print(
      ggplot(tac_df, aes(x = frame_mid, y = .data[[reg]])) +
        geom_line(color = "steelblue", linewidth = 1) +
        geom_point(color = "darkred",  size = 2) +
        labs(
          title = sprintf("TAC – %s (sub %s, ses %s, %s)",
                          reg, sub, ses, cond),
          x = "Time (min)",
          y = "Radioactivity (kBq/mL)"
        ) +
        theme_bw(base_size = 14)
    )
    dev.off()
    message("Saved: ", outpath)
  }
}

#---------------------------------------------------------------------
# 2) Mean ± SD TAC per region across all scans
#---------------------------------------------------------------------

# Unnest to: frame_mid × Region × TAC, keeping subject/session/condition
all_tacs <- modeldata %>%
  select(sub, ses, condition, tacdata) %>%
  unnest(tacdata) %>%
  pivot_longer(
    cols      = all_of(regions_vector),
    names_to  = "Region",
    values_to = "TAC"
  )

mean_tacs <- all_tacs %>%
  group_by(Region, frame_mid) %>%
  summarise(
    mean_TAC = mean(TAC, na.rm = TRUE),
    sd_TAC   = sd(TAC,   na.rm = TRUE),
    .groups  = "drop"
  )

for (reg in regions_vector) {
  outdir <- file.path(kinfitr_output_dir, "02_Outcomes", "01_TAC_plots", reg)
  dir.create(outdir, recursive = TRUE, showWarnings = FALSE)

  outpath <- file.path(outdir,
                       sprintf("Region-%s_mean_TAC.png", reg))

  df <- filter(mean_tacs, Region == reg)

  png(outpath, width = 1000, height = 800)
  print(
    ggplot(df, aes(x = frame_mid, y = mean_TAC)) +
      geom_ribbon(aes(ymin = mean_TAC - sd_TAC,
                      ymax = mean_TAC + sd_TAC),
                  fill = "grey80", alpha = 0.5) +
      geom_line(color = "steelblue", linewidth = 1.2) +
      geom_point(color = "darkred", size = 2) +
      labs(
        title = paste0("Average TAC – ", reg),
        x = "Time (min)",
        y = "Mean radioactivity (kBq/mL)"
      ) +
      theme_bw(base_size = 14)
  )
  dev.off()
  message("Saved mean plot: ", outpath)
}
```


## Save the environment variables to use in later scripts
```{r}
# Define folder for processed RDS files
processed_data_dir <- file.path(kinfitr_output_dir, "01_processed_data")
dir.create(processed_data_dir, showWarnings = FALSE)

# Save key objects as RDS
saveRDS(idif_data, file = file.path(processed_data_dir, "idif_data.rds"))
saveRDS(glc_data, file = file.path(processed_data_dir, "glc_data.rds"))
saveRDS(volume_weighted_average_tacs, file = file.path(processed_data_dir, "volume_weighted_average_tacs"))
saveRDS(modeldata, file = file.path(processed_data_dir, "modeldata.rds"))

# Optional intermediates
saveRDS(selected_tacs, file = file.path(processed_data_dir, "selected_tacs.rds"))
saveRDS(regions_vector, file = file.path(processed_data_dir, "regions_vector.rds"))

```
