---
title: "HDP_pet_modeling_patlak"
output: html_notebook
date: "2025-12-09"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Setup

## Packages

First we load up the packages we'll be working with

```{r}
# remotes::install_github("mathesong/bloodstream")
# remotes::install_github("mathesong/kinfitr")

rm(list = ls())  # Clear all objects from the workspace

library(tidyverse)
library(bloodstream)
library(kinfitr)
library(job)
library(dplyr)
library(tidyr)
library(readr)
library(ggpubr)
library(jsonlite)
library(lme4)
library(emmeans)
library(lmerTest)

theme_set(theme_light())


```

# Loading Data

```{r}

find_repo_root <- function(start = getwd()) {
  current <- normalizePath(start, winslash = "/", mustWork = TRUE)
  repeat {
    if (file.exists(file.path(current, "config", "paths.json"))) {
      return(current)
    }
    parent <- dirname(current)
    if (identical(parent, current)) {
      stop("Could not locate repo root containing config/paths.json")
    }
    current <- parent
  }
}
repo_root <- find_repo_root(getwd())
source(file.path(repo_root, "code", "utils", "project_paths.R"))
paths <- load_project_paths(repo_root)
bids_dir <- paths$bids_root
kinfitr_output_dir <- file.path(paths$kinfitr_output_root)

# Define the paths relative to the base path
idif_path <- file.path(bids_dir, "derivatives/idif_aorta")
participants_file <- file.path(bids_dir, "participants.tsv")


# Read the participants.tsv file
participants <- read.delim(participants_file)

processed_data_dir <- file.path(kinfitr_output_dir, "01_processed_data")

idif_data      <- readRDS(file.path(processed_data_dir, "idif_data.rds"))
glc_data       <- readRDS(file.path(processed_data_dir, "glc_data.rds"))
volume_weighted_average_tacs  <- readRDS(file.path(processed_data_dir, "volume_weighted_average_tacs"))
modeldata      <- readRDS(file.path(processed_data_dir, "modeldata.rds"))

# Optional
selected_tacs  <- readRDS(file.path(processed_data_dir, "selected_tacs.rds"))
regions_vector <- readRDS(file.path(processed_data_dir, "regions_vector.rds"))
```


## Fitting multiple TACs

### Selecting a t\* value

Let's choose an appropriate t\* value

```{r, fig.width=12, fig.height=12}
modeldata %>% 
  ungroup() %>% 
  #filter(condition=="Placebo") %>% 
  slice(1:10) %>% 
  mutate(tstarplot = map2(tacdata, input, 
     ~Patlak_tstar(
         t_tac = .x$frame_mid, 
         lowroi =  .x$hippocampus,
         medroi =  .x$insula,
         highroi = .x$frontal_cortex, 
         input = .y)#,
         #vB = 0.05)
     )) %>% 
  pull(tstarplot)
```


## Now do the Patlak modeling:

```{r}

modeldata_long <- modeldata %>%
     unnest(tacdata) %>%
     select(-frame_start, -frame_end, -frame_dur, -meanTAC)

modeldata_long <- modeldata_long %>%
     gather(key = Region, value = TAC, -weights, 
            -ses, -sub, -task, -trc, -acq, -run, -rec, -condition,
            -input_idif, -input, -input_feng, -frame_mid)

modeldata_long <- modeldata_long %>%
     group_by(ses, sub, Region, task, trc, acq, run, rec, condition, input_idif, input) %>%
     nest(tacdata = -c(ses, sub, Region, task, trc, acq, run, rec, condition, input_idif, input, input_feng))

modeldata_long <- modeldata_long %>% 
  group_by(sub, condition, Region) %>% 
  mutate(Patlak_fit = map2(tacdata, input_feng, ~Patlakplot(t_tac = .x$frame_mid,
                                             tac = .x$TAC, 
                                             input= .y, 
                                             tstarIncludedFrames = 10, 
                                             weights = .x$weights)))#, 
                                             #vB = 0.05)))

saveRDS(modeldata_long, file.path(kinfitr_output_dir, "01_processed_data", "modeldata_Patlak.rds"))


```

## Save the workspace after modeling
```{r}
dir.create(file.path(kinfitr_output_dir, "workspaces"), recursive = TRUE, showWarnings = FALSE)
save.image(file = file.path(kinfitr_output_dir, "workspaces", "02a_HDP_pet_modeling_Patlak_workspace.RData"))  # Save entire workspace to file```
```

## Load workspace to start the script from here
```{r}
load(file.path(kinfitr_output_dir, "workspaces", "02a_HDP_pet_modeling_Patlak_workspace.RData"))
```

### Plotting

```{r}
library(purrr)
library(fs)

# Define model types to loop over
model_types <- c("Patlak_fit")

for (model_type in model_types) {
  
  message("Processing model: ", model_type)
  
  for (i in seq_len(nrow(modeldata_long))) {
    
    sub  <- modeldata_long$sub[i]
    ses  <- modeldata_long$ses[i]
    cond <- modeldata_long$condition[i]
    reg  <- modeldata_long$Region[i]
    
    # Create a model-specific output folder
    outdir <- file.path(kinfitr_output_dir, "02_Outcomes", "02_model_fits", model_type, reg)
    dir_create(outdir, recurse = TRUE)
    
    # Make a clean filename
    fname <- sprintf("sub-%s_ses-%s_%s_%s_%s.png",
                     sub, ses, cond, reg, model_type)
    
    outpath <- file.path(outdir, fname)
    
    # Open plotting device
    png(outpath, width = 1000, height = 800)
    
    # Plot the model fit
    fit_obj <- modeldata_long[[model_type]][[i]]
    if (!is.null(fit_obj)) {
      print(plot(fit_obj))
    }
    
    dev.off()
    
    message("Saved: ", outpath)
  }
}

```
```{r fig.width=12, fig.height=12}

# ================================================================
# --- Helper function to extract outcomes from Logan / Patlak fits
# ================================================================

LC = 0.65 # Lumped Constant for k4=0 from Wu et al. 2003

extract_linear_outcomes <- function(data, fit_col, model_label, glc_data = NULL) {
  
  df <- data %>%
    select(sub, condition, ses, Region, !!sym(fit_col))
  
  if (fit_col == "Patlak_fit") {
    df <- df %>%
      mutate(
        K = map_dbl(!!sym(fit_col), c("par", "K")),
        Model = "Patlak"
      )
    
    # If glucose data available, compute CMRglc
    if (!is.null(glc_data)) {
      df <- df %>%
        left_join(glc_data %>% select(sub, condition, ses, glc_conc),
                  by = c("sub", "condition", "ses")) %>%
        mutate(CMRglc = K * glc_conc * 100 / LC) %>% # convert to typical units of CMRglc: µmol/100g/min
        select(-glc_conc)
    }
  }
    
  # Remove list-column and reorder columns (Model first)
  df <- df %>%
    select(Model, everything(), -!!sym(fit_col))
  
  return(df)
}


# --- Apply extraction for both models
Patlak_outcomes_all <-
  extract_linear_outcomes(modeldata_long, "Patlak_fit", "Patlak", glc_data)


# --- Save results
dir.create(
  file.path(kinfitr_output_dir, "02_Outcomes", "03_model_outcomes"),
  recursive = TRUE,
  showWarnings = FALSE
)

write_tsv(Patlak_outcomes_all,
          file.path(kinfitr_output_dir, "02_Outcomes", "03_model_outcomes", "Patlak_outcomes.tsv"))


# --- Helper function for plotting
plot_and_save <- function(df, metric, ylabel, title, filename) {
  
  p <- ggplot(df, aes(x = condition, y = .data[[metric]], colour = sub, group = sub)) +
    geom_point(size = 3, colour = "black") +
    geom_point(size = 2.5) +
    geom_line() +
    facet_wrap(~Region, scales = "free_y") +
    labs(x = "Condition", y = ylabel, title = title) +
    theme_bw(base_size = 13) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      strip.text = element_text(size = 12)
    )
  
  out_dir <- file.path(kinfitr_output_dir, "02_Outcomes", "03_model_outcomes", "plots")
  dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)
  
  ggsave(
    filename = file.path(out_dir, paste0(filename, ".png")),
    plot = p, width = 12, height = 8
  )
  
  print(p)
}


# --- Plot metrics for each model

# Patlak metrics
metrics_patlak <- c("K", "CMRglc")
titles_patlak  <- c("Ki", "CMRglc")

walk2(metrics_patlak, titles_patlak,
      ~ plot_and_save(Patlak_outcomes_all,
                      .x, .x,
                      paste("Comparison of", .y, "by Condition and Region"),
                      paste0(.x, "_Patlak")))

```


```{r}
library(effectsize)

# --- CMRglc values ---

# Filter data for Placebo and Verum (CMRglc values) for Whole Brain
placebo_CMRglc <- Patlak_outcomes_all %>%
  filter(condition == "Placebo", Region == "whole_brain") %>%
  arrange(sub) %>%
  pull(CMRglc)

verum_CMRglc <- Patlak_outcomes_all %>%
  filter(condition == "Verum", Region == "whole_brain") %>%
  arrange(sub) %>%
  pull(CMRglc)

# Perform a paired t-test
t_test_result_CMRglc <- t.test(verum_CMRglc, placebo_CMRglc, paired = TRUE, alternative = 'greater', conf.level = 0.95)

# Calculate means and standard deviations
mean_placebo_CMRglc <- mean(placebo_CMRglc)
sd_placebo_CMRglc <- sd(placebo_CMRglc)

mean_verum_CMRglc <- mean(verum_CMRglc)
sd_verum_CMRglc <- sd(verum_CMRglc)

# Calculate mean % increase from placebo to verum
percent_increase_CMRglc <- ((mean_verum_CMRglc - mean_placebo_CMRglc) / mean_placebo_CMRglc) * 100

# Calculate Cohen's d for paired samples
cohen_d_CMRglc <- repeated_measures_d(t_test_result_CMRglc, adjust = TRUE, method="z")

# Print results
print(t_test_result_CMRglc)
cat("\nMeans and Standard Deviations (CMRglc values):\n")
cat("Placebo mean:", mean_placebo_CMRglc, "SD:", sd_placebo_CMRglc, "\n")
cat("Verum mean:", mean_verum_CMRglc, "SD:", sd_verum_CMRglc, "\n")
cat("Mean % increase from Placebo to Verum:", percent_increase_CMRglc, "%\n")
cat("95% CI of difference:", t_test_result_CMRglc$conf.int[1], "to", t_test_result_CMRglc$conf.int[2], "\n")
cat("Cohen's d (paired):", cohen_d_CMRglc$d, "\n")


```



```{r}
# -------------------------------------------------------------
# Session/order effect within Placebo only (whole_brain)
# -------------------------------------------------------------

# Subset placebo, whole_brain
placebo_data <- Patlak_outcomes_all %>%
  filter(condition == "Placebo", Region == "whole_brain")

# Split by session (placebo in session 01 vs 02)
placebo_ses01 <- placebo_data %>%
  filter(ses == "01") %>%
  arrange(sub)

placebo_ses02 <- placebo_data %>%
  filter(ses == "02") %>%
  arrange(sub)

# --- CMRglc: placebo session-order comparison (independent groups) ---

CMR_ses01 <- placebo_ses01$CMRglc
CMR_ses02 <- placebo_ses02$CMRglc

t_test_placebo_CMR <- t.test(CMR_ses01, CMR_ses02,
                             paired = FALSE,
                             alternative = "two.sided",
                             conf.level = 0.95)

mean_CMR_ses01 <- mean(CMR_ses01, na.rm = TRUE)
sd_CMR_ses01   <- sd(CMR_ses01,   na.rm = TRUE)

mean_CMR_ses02 <- mean(CMR_ses02, na.rm = TRUE)
sd_CMR_ses02   <- sd(CMR_ses02,   na.rm = TRUE)

# Optional: Cohen's d for independent samples
# cohen_d_placebo_CMR <- effsize::cohen.d(CMR_ses01, CMR_ses02, hedges.correction = TRUE)

cat("\n--- Placebo only: Session-order effect on CMRglc (whole_brain) ---\n")
print(t_test_placebo_CMR)
cat("\nMeans and SDs (CMRglc):\n")
cat("Placebo in ses-01: mean =", mean_CMR_ses01, "SD =", sd_CMR_ses01, "\n")
cat("Placebo in ses-02: mean =", mean_CMR_ses02, "SD =", sd_CMR_ses02, "\n")
cat("95% CI of difference (ses-01 - ses-02):",
    t_test_placebo_CMR$conf.int[1], "to", t_test_placebo_CMR$conf.int[2], "\n")
# cat("Cohen's d (independent):", cohen_d_placebo_CMR$estimate, "\n")
```


```{r}
# --- Session-adjusted repeated-measures model for whole-brain CMRglc ----------
# This tests the condition effect while adjusting for whether the scan was
# acquired in session 1 or session 2.

whole_brain_long <- Patlak_outcomes_all %>%
  filter(Region == "whole_brain", condition %in% c("Placebo", "Verum")) %>%
  mutate(
    ses = case_when(
      as.character(ses) %in% c("ses-01", "01", "1") ~ "ses-01",
      as.character(ses) %in% c("ses-02", "02", "2") ~ "ses-02",
      TRUE ~ NA_character_
    ),
    condition = factor(condition, levels = c("Verum", "Placebo")),
    ses = factor(ses, levels = c("ses-01", "ses-02"))
  ) %>%
  drop_na(CMRglc, condition, ses, sub)

session_adjusted_model_CMRglc <- lmer(
  CMRglc ~ condition + ses + (1 | sub),
  data = whole_brain_long
)

cat("\nSession-adjusted mixed model for whole-brain CMRglc:\n")
print(summary(session_adjusted_model_CMRglc))

emm_condition <- emmeans(session_adjusted_model_CMRglc, ~ condition)

cat("\nEstimated marginal means for condition (adjusted for session):\n")
print(emm_condition)

cat("\nAdjusted Verum vs Placebo contrast:\n")
print(pairs(emm_condition))
```


```{r}

library(dplyr)
library(lme4)
library(emmeans)
library(purrr)
library(tibble)

# ----------------------------
# Define subcortical regions
# ----------------------------
regions <- unique(Patlak_outcomes_all$Region)

subcortical_regions_expected <- c(
  "insula", "striatum", "hippocampus", "amygdala", "thalamus",
  "caudate_nucleus", "putamen", "pallidum", "nucleus_accumbens",
  "choroid_plexus", "ventral_diencephalon"
)

subcortical_regions <- intersect(subcortical_regions_expected, regions)
missing_subcortical_regions <- setdiff(subcortical_regions_expected, regions)

if (length(missing_subcortical_regions) > 0) {
  warning(
    "The following expected subcortical regions are not present in Patlak_outcomes_all: ",
    paste(missing_subcortical_regions, collapse = ", ")
  )
}

# Conditions to compare (edit if needed)
cond_a <- "Verum"
cond_b <- "Placebo"

# ----------------------------
# Region-wise paired t-test summary table
# ----------------------------
subcortical_table <- map_dfr(subcortical_regions, function(reg) {

  wide <- Patlak_outcomes_all %>%
    filter(Region == reg, condition %in% c(cond_a, cond_b)) %>%
    select(sub, condition, CMRglc) %>%
    distinct() %>%
    pivot_wider(names_from = condition, values_from = CMRglc)

  # keep only complete pairs
  wide <- wide %>% filter(!is.na(.data[[cond_a]]) & !is.na(.data[[cond_b]]))

  verum   <- wide[[cond_a]]
  placebo <- wide[[cond_b]]
  d <- verum - placebo

  tt <- t.test(verum, placebo, paired = TRUE)#, alternative = "greater")

  mean_placebo <- mean(placebo)
  mean_verum   <- mean(verum)
  mean_diff    <- mean(d)
  percent_inc  <- ((mean_verum - mean_placebo) / mean_placebo) * 100

  # paired effect size (Cohen's dz)
  cohen_dz <- mean_diff / sd(d)

  tibble(
    Region = reg,
    n_pairs = length(d),
    mean_placebo = mean_placebo,
    mean_verum = mean_verum,
    mean_diff = mean_diff,
    percent_increase = percent_inc,
    t = unname(tt$statistic),
    df = unname(tt$parameter),
    p_value = unname(tt$p.value),
    ci_low_mean_diff  = unname(tt$conf.int[1]),
    ci_high_mean_diff = unname(tt$conf.int[2]),
    cohen_dz = cohen_dz
  )
}) %>%
  mutate(
    p_FDR = p.adjust(p_value, method = "fdr")
  ) %>%
  select(
    Region, n_pairs,
    mean_placebo, mean_verum, mean_diff, percent_increase,
    t, df, p_value, p_FDR,
    ci_low_mean_diff, ci_high_mean_diff,
    cohen_dz
  )

# Output table
subcortical_table
write.csv(subcortical_table, file.path(kinfitr_output_dir, "02_Outcomes", "03_model_outcomes", "subcortical_paired_ttest_results.csv"), row.names = FALSE)
```




